### 2017-02-16 Paper_v1.tex
Feb16

To better understand which parameters and parameter combinations influenced the performance of the kinetic model, we performed sensitivity analysis (Fig.~\ref{fig:Sensitivity}).
We perturbed each $V^{max}$ parameter, either individually or in pairwise combinations and measured the change in either CAT production or the overall system state.
CAT production was most sensitive to the CAT synthesis reaction, oxidative phosphorylation, and the pyruvate-consuming alanine synthesis reaction (Fig.~\ref{fig:Sensitivity}, top, section A).
%Taking into account these three reactions, and the next 16 most important reactions (section B), we saw a common theme of reactions involving the cofactors ATP, NADH, NADPH, and coenzyme A, as well as the metabolites pyruvate and glutamate.
We saw a common theme of the most important reactions producing or consuming the cofactors ATP, NADH, NADPH, and coenzyme A, as well as the metabolites pyruvate and glutamate.
Of the 25 reactions to which CAT production was most sensitive, 9 produced or consumed ATP, making it the most represented in these top reactions (with the exception of hydrogen and phosphate ion).
The next most represented were pyruvate, glutamate, and ADP with 7 reactions each, followed by coenzyme A, $\alpha$-ketoglutarate, NAD/NADH, and NADP/NADPH, with 6 reactions each.
This makes sense, as glutamate was an important precursor for the synthesis of other amino acids required by CAT production.
Meanwhile, the cofactors provided energy to power CAT synthesis, while pyruvate was important for energy generation following glucose depletion.
In addition, pyruvate was required for the synthesis of several amino acids.
The pairwise sensitivities (off-diagonal elements) were different from the corresponding first-order sensitivities (diagonal elements), and led to interesting outcomes.
The combination of certain reactions had a much greater or lesser effect on CAT production than that of the individual reactions by themselves.
For example, glutamine synthesis and arginine degradation were both among the most important reactions to CAT production (they ranked 5th and 10th, respectively).
This was likely because they both affected the sensitive glutamine-glutamate balance; glutamine synthesis consumes glutamate, while arginine degradation produces it.
However, when both were perturbed, their combined effect on the model was low, as the respective contributions to consumption and production of glutamate cancelled.
An example of positive synergy can be seen in cometabolite interconversion.
Pyridine nucleotide transhydrogenase catalyzes two reactions: one converts NAD and NADPH into NADH and NADP, while the other does the reverse and also generates a proton gradient.
Increasing one or the other has little effect on the model, as any increase in reaction flux is hampered by the imbalance created.
Increasing both, however, allows all cometabolites to balance, so that the reaction fluxes increase unhindered.
Thus, the pairwise sensitivity is much higher than the sum of the first-order sensitivities.

The overall system state was also sensitive to cofactors and substrates; however, instead of pyruvate and glutamate, the substrates driving metabolism were pyruvate and G3P.
The system was most sensitive to an oxidative phosphorylation reaction (cytochrome oxidase), which converted ubiquinol to ubiquinone while generating a proton gradient.
The next 4 most important reactions were all consumers of pyruvate: lactate dehydrogenase, pyruvate formate lyase, alanine synthesis, and PEP synthase.
Of the 25 reactions to which the system was most sensitive, 7 produced or consumed pyruvate and 7 participated in NAD/NADH exchange.
With the exception of hydrogen, these were the most represented species in the top 25 reactions.
Also important were coenzyme A (5 reactions) and acetyl coenzyme A (4 reactions), as well as ubiquinone/ubiquinol, NADP/NADPH, G3P, and ATP  (4 reactions each).
%We saw a common theme of the most important reactions producing or consuming the cofactors ATP, NADH, NADPH, ubiquinone/ubiquinol, and coenzyme A, as well as the metabolites pyruvate and G3P.
%These two and the next 30 most important reactions (section G) largely involved cofactors, especially ATP and NADPH, as well as substrate-consuming reactions and oxidative phosphorylation.
The system state also had pairwise sensitivities that differed from the corresponding first-order sensitivities and stood out as significant.
For example, alanine degradation was among the most important reactions, as it produced pyruvate; GMP synthesis was also moderately important, as it produced glutamate.
However, when both reactions were increased, the combined effect on the model was almost zero.
%This can be understood by also considering a third reaction: alanine synthesis from pyruvate and glutamate.
This can be understood by considering the reactions that involve both pyruvate and glutamate: they all either produced both of these substrates or consumed both.
When alanine degradation was increased, the excess pyruvate stimulated these reactions to consume both pyruvate and glutamate; the amounts of both of these substrates could not be conserved.
But when GMP synthesis was perturbed as well, the glutamate deficiency was corrected, cancelling much of the effect on the system.
One of the pyruvate- and glutamate-consuming reactions was alanine synthesis; in this case, the levels of pyruvate, glutamate, and alanine were all virtually unchanged.
%An example of positive synergy can be seen in the interplay of amino acid synthesis and the TCA cycle: aspartate synthesis and the reverse reaction of succinyl coenzyme A synthetase are both relatively unimportant to the system.
An example of positive synergy can be seen in histidine synthesis, one of the least influential reactions that consumes three units of ATP.
When perturbed in combination with the reverse reaction of succinyl coenzyme A synthetase, another reaction with little overall effect on the model, the combined effect on the system is much greater.
This may be because the reverse reaction of succinyl coenzyme A synthetase produced ATP, which further stimulated histidine synthesis.
%Another interesting result was the intersection of sections F and G with section J.
%The 53 reactions in section J were turned off in the best-fit set ($V^{max}=0$);
%therefore, the perturbation of these reactions had no effect on the model.
%As a result, all pairwise sensitivities with reactions in section J were pseudo first-order sensitivities for the other reactions.
%Interestingly, many reactions in section F and several in section G showed their highest sensitivities when paired with the "non-effects" of section J.
%\textcolor{red}{NICK: DOES THIS MAKE SENSE? WHY ARE THESE DIFFERENT THAN THE SINGLE PERTURBATIONS?}
%Of these, three involved pyruvate, strengthening its role as a key metabolite; the others were glucose consumption via GTP/CTP-specific hexokinases,
%fumarate reductase, and SO$_4$ utilization.
%This suggested that these reactions' effects on the model were canceled out or lessened by most other reactions, but were of course not affected by the reactions in section J.
%This was also likely the reason that section J ranked above section K, despite having no effect on the model themselves.
Taken together, sensitivity analysis identified blocks of parameters that either individually, or in combination influenced model performance.

To understand whether the CFPS performance was optimal, we calculated the carbon yield and energy efficiency of CAT production (Fig.~\ref{CAT_balances}).
The best-fit parameter set for the kinetic model predicted a CAT carbon yield of 7.9\%, while the experimental dataset had a CAT carbon yield of 8.2\%.
This was calculated as the increase in CAT concentration times the CAT carbon number, divided by the sum of the consumption terms for glucose and all amino acids excpet arginine and glutamate, as no data were available for these, weighted by their respective carbon numbers.
To explain where the remainder of carbon was going, we performed a carbon balance for the best-fit set (Fig.~\ref{fig:CAT_balances}A).
Of the other 92\% of carbon, 35\% accumulated as organic acids (lactate, acetate, succinate and malate) and 9\% accumulated as amino acids (alanine and glutamine).
The remaining 48\% went to the net accumulation of all other metabolites, particularly carbon dioxide.
The best-fit set and the experimental dataset both produced CAT with an energy efficiency of 7\% (Fig.~\ref{fig:CAT_balances}B).
This was calculated as the increase in CAT concentration times the CAT number of equivalent ATP molecules, divided by glucose consumption times times the number of equivalent ATP molecules for glucose, equal to 15 in the optimal case.
%An additional 62\% of energy went to the accumulation of glycolysis metabolites, 6\% to TCA cycle metabolites, and 25\% to acetate and lactate.
An additional 62\% of energy went to the accumulation of glycolysis metabolites, and 31\% to organic acids (lactate, acetate, succinate and malate).
This shows that there is much room for improvement of the efficiency of CFPS.
Gene knockouts in the electron transport chain further reduced the performance of the CFPS reaction (Fig.~\ref{fig:oxKO_ssFBA_yields}).
A key finding of both the CAT and overall system state sensitivity analysis was the importance of oxidative phosphorylation.
To investigate this further, we knocked out key oxidative phosphorylation reactions in the ensemble of kinetic models to examine the effect on CAT production and carbon yield.
A single \textit{cyd} knockout reduced the CAT carbon yield from 7.9\% to 2.6\% (Table~\ref{tbl:yield_breakdown}).
On the other hand, a \textit{nuo} knockout showed a less dramatic decrease in yield,
reducing the CAT carbon yield to 6.9\%.
Knocking out \textit{app} increased CAT yield to 8.1\%, but this increase was not statistically significantly different from that of the control.
Lastly, knocking out all three reactions reduced the CAT yield to 0.7\%, similar to knocking out the \textit{cyd} alone.
Thus, the model suggested the key step in oxidative phosphorylation was catalyzed by the gene product of \textit{cyd}.
However, while disruption of \textit{cyd} significantly reduced the CAT carbon yield, it did not completely eliminate the production of CAT.
This suggested there was a mixture of energy sources supporting CAT production, with the most significant being oxidative phosphorylation.
%The \textit{cyd} enzyme is responsible for transferring electrons with oxygen as an electron acceptor allowing the release of energy to form ATP.

@@

% This array of sensitivity indices was also organized into a clustergram (Fig.~\ref{fig:Sensitivity}, bottom).
% Theoretical maximum CAT carbon yields for three cases discussed below were calculated using flux balance anaylsis (FBA) with a sequence-specific description of CAT synthesis, where $\Delta m_{i}$ denotes the flux of the $i^{th}$ species.
% The transcription and translation template reactions come from sequence-specific analysis \cite{Allen:2003aa}, and include transcription initiation, transcription, mRNA degradation, translation initiation, translation, and tRNA charging.
% The rest of the network followed a pseudo steady-state assumption where metabolites were not allowed to accumulate; thus, the network could be solved by linear programming instead of solving differential equations.
% \begin{eqnarray*}
% G+R_{1} &\longrightarrow& G^{*}\\
% G^{*}+151 \cdot gtp+144 \cdot ctp+189 \cdot utp+176 \cdot atp &\longrightarrow& mRNA+G+R_{1}+1320 \cdot Pi\\
% mRNA &\longrightarrow& 151 \cdot gmp+144 \cdot cmp+189 \cdot ump+176 \cdot amp\\
% \end{eqnarray*}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=0.76\textwidth,trim=0 28 615 40,clip]{./Figures/Sensitivity.pdf}
\caption{Normalized first-order and pairwise sensitivities of CAT production (top) and system state (bottom) to maximum reaction rates.}
\label{fig:Sensitivity}
\end{figure}

%\begin{figure}[ht]
%\centering
%\includegraphics[width=1\textwidth]{./Figures/oxphos_ko.pdf}
%\caption{The effect of oxidative phosphorylation on glucose uptake, CAT production and CAT carbon yield. A. 95\% confidence interval of an ensemble for glucose concentration versus time for no knockouts (blue shaded region), \textit{cyd} knockout (green shaded region), and \textit{nuo} knockout (red shaded region). B. 95\% confidence interval of an ensemble for CAT concentration versus time for no knockouts (blue shaded region), \textit{cyd} knockout (green shaded region), and \textit{nuo} knockout (red shaded region). C. CAT carbon yield for 5 different cases of oxidative phosphorylation: no knockouts (blue), \textit{cyd} knockout (green), \textit{nuo} knockout (red), \textit{app} knockout (light gray), and a combination of \textit{cyd}, \textit{nuo}, \textit{app} knockouts (dark gray).}
%\label{fig:oxphos_ko}
%\end{figure}
%\clearpage

%\begin{figure}[ht]
%\centering
%\includegraphics[width=1\textwidth]{./Figures/CAT_prod_yield.pdf}
%\caption{Sequence-specific flux balance analysis of CAT production and yield. A. 95\% confidence interval of the ensemble (light blue region) for CAT concentration versus time. B. Theoretical maximum carbon yield of CAT calcualted by ssFBA for three different cases: constrained by transcription/translation (TX/TL) rates (black), same as previous but without amino acid synthesis reactions (gray), and same as previous but constrained by experimental measurements where available (light gray). Error bars represent standard deviation of the ensemble.}
%\label{fig:CATProdYield}
%\end{figure}
%\clearpage

@@

>>>>

To better understand which parameters and parameter combinations influenced the performance of the kinetic model, we performed sensitivity analysis (Fig.~\ref{fig:Sensitivity-CAT}).
We perturbed each $V^{max}$ parameter, either individually or in pairwise combinations and measured the change in either CAT production or the overall system state.
The eigen decomposition of the sensitivity shows that CAT synthesis and oxidative phosphorylation are the most important to overall CAT production, followed by the pyruvate-consuming alanine synthesis reaction.
Among the top 20 reactions, we saw a common theme of the cofactors ATP, NADH, NADPH, and coenzyme A, as well as the metabolites pyruvate, glutamate, and $\alpha$-ketoglutarate.
ATP appears 7 times in these 20 reactions, while pyruvate, glutamate, and coenzyme A appear 6 times each.
NADH, NADPH, and $\alpha$-ketoglutarate each appear 5 times.
This result makes sense, as the high energy cost of protein synthesis means that energy cofactors played a crucial role.
Also, pyruvate served as the primary substrate after glucose ran out, and pyruvate, glutamate, and $\alpha$-ketoglutarate were all important precursors for the synthesis of amino acids required by CAT production.
%Meanwhile, the cofactors provided energy to power CAT synthesis, while pyruvate was important for energy generation following glucose depletion.
We performed the same eigen decomposition on the sensitivity of the overall system state to network reactions (Fig.~\ref{fig:Sensitivity-System}).
Cytochrome oxidase, part of oxidative phosphorylation, was seen to have the greatest effect on the system state.
Next most influential was the forward reaction of lactate dehydrogenase, followed by NADH:ubiquinone oxidoreductase, another oxidative phosphorylation reaction.
The overall system state was also sensitive to cofactors and substrates, specifically NADH and pyruvate.
%The system was most sensitive to an oxidative phosphorylation reaction (cytochrome oxidase), which converted ubiquinol to ubiquinone while generating a proton gradient.
%The next 4 most important reactions were all consumers of pyruvate: lactate dehydrogenase, pyruvate formate lyase, alanine synthesis, and PEP synthase.
Among the top 20 reactions, NADH appears in 8 reactions, pyruvate in 6, coenzyme A in 5, and ATP in 4.
Glutamate, $\alpha$-ketoglutarate, G3P, and ubiquinone/ubiquinol appear 3 times each.
Taken together, sensitivity analysis identified that substrates and energy cofactors, specifically those around oxidative phosphorylation, most influenced model performance.

To understand whether the CFPS performance was optimal, we calculated the carbon yield and energy efficiency of CAT production (Fig.~\ref{fig:CAT_balances}).
The best-fit parameter set for the kinetic model predicted a CAT carbon yield of 7.9\%, while the experimental dataset had a CAT carbon yield of 8.2\%.
This was calculated as the increase in CAT concentration times the CAT carbon number, divided by the sum of the consumption terms for glucose and all amino acids excpet arginine and glutamate, as no data were available for these, weighted by their respective carbon numbers.
To explain where the remainder of carbon was going, we performed a carbon balance for the best-fit set (Fig.~\ref{fig:CAT_balances}A).
Of the other 92\% of carbon, 35\% accumulated as organic acids (lactate, acetate, succinate and malate) and 9\% accumulated as amino acids (alanine and glutamine).
The remaining 48\% went to the net accumulation of all other metabolites, particularly carbon dioxide.
The best-fit set and the experimental dataset both produced CAT with an energy efficiency of 7\% (Fig.~\ref{fig:CAT_balances}B).
This was calculated as the increase in CAT concentration times the CAT number of equivalent ATP molecules, divided by glucose consumption times times the number of equivalent ATP molecules for glucose, equal to 15 in the optimal case.
%An additional 62\% of energy went to the accumulation of glycolysis metabolites, 6\% to TCA cycle metabolites, and 25\% to acetate and lactate.
An additional 62\% of energy went to the accumulation of glycolysis metabolites, and 31\% to organic acids (lactate, acetate, succinate and malate).
This shows that there is much room for improvement of the efficiency of CFPS.
Gene knockouts in the electron transport chain further reduced the performance of the CFPS reaction (Fig.~\ref{fig:oxKO_ssFBA_yields}).
A key finding of both the CAT and overall system state sensitivity analysis was the importance of oxidative phosphorylation.
To investigate this further, we knocked out key oxidative phosphorylation reactions in the ensemble of kinetic models to examine the effect on CAT production and carbon yield.
A single \textit{cyd} knockout reduced the CAT carbon yield from 7.9\% to 2.6\% (Table~\ref{tbl:yield_breakdown}).
On the other hand, a \textit{nuo} knockout showed a less dramatic decrease in yield,
reducing the CAT carbon yield to 6.9\%.
Knocking out \textit{app} increased CAT yield to 8.1\%, but this increase was not statistically significantly different from that of the control.
Lastly, knocking out all three reactions reduced the CAT yield to 0.7\%, similar to knocking out the \textit{cyd} alone.
Thus, the model suggested the key step in oxidative phosphorylation was catalyzed by the gene product of \textit{cyd}.
However, while disruption of \textit{cyd} significantly reduced the CAT carbon yield, it did not completely eliminate the production of CAT.
This suggested there was a mixture of energy sources supporting CAT production, with the most significant being oxidative phosphorylation.
%The \textit{cyd} enzyme is responsible for transferring electrons with oxygen as an electron acceptor allowing the release of energy to form ATP.

@@

% This array of sensitivity indices was also organized into a clustergram (Fig.~\ref{fig:Sensitivity-CAT}, bottom).
% Theoretical maximum CAT carbon yields for three cases discussed below were calculated using flux balance anaylsis (FBA) with a sequence-specific description of CAT synthesis, where $\Delta m_{i}$ denotes the flux of the $i^{th}$ species.
% The transcription and translation template reactions come from sequence-specific analysis \cite{Allen:2003aa}, and include transcription initiation, transcription, mRNA degradation, translation initiation, translation, and tRNA charging.
% The rest of the network followed a pseudo steady-state assumption where metabolites were not allowed to accumulate; thus, the network could be solved by linear programming instead of solving differential equations.
% \begin{eqnarray*}
% G+R_{1} &\longrightarrow& G^{*}\\
% G^{*}+151 \cdot gtp+144 \cdot ctp+189 \cdot utp+176 \cdot atp &\longrightarrow& mRNA+G+R_{1}+1320 \cdot Pi\\
% mRNA &\longrightarrow& 151 \cdot gmp+144 \cdot cmp+189 \cdot ump+176 \cdot amp\\
% \end{eqnarray*}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=0.95\textwidth]{./Figures/Sensitivity-CAT.pdf}
\caption{Sensitivity of CAT production to model parameters, decomposed into eigenmodes. Sensitivity of top 10 eigenvectors  (columns, left to right) to top 20 most influential parameters (rows, bottom to top). Relative magnitudes of eigenvalues expressed as percentages.}
\label{fig:Sensitivity-CAT}
\end{figure}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=0.9\textwidth]{./Figures/Sensitivity-System.pdf}
\caption{Sensitivity of system state to model parameters, decomposed into eigenmodes. Sensitivity of top 10 eigenvectors  (columns, left to right) to top 20 most influential parameters (rows, bottom to top). Relative magnitudes of eigenvalues expressed as percentages.}
\label{fig:Sensitivity-System}
\end{figure}
